---
title: "P8105 Homework 3"
author: "Shehzrin Shah"
date: "2024-10-12"
output: github_document
---

```{r setup, echo = FALSE, message = FALSE}
library(tidyverse)
library(p8105.datasets)
library(ggridges)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Problem 1 

Loading the data from the `p8105.datasets` package: 
```{r}
data("ny_noaa")
summary(ny_noaa)
```

The dataset being explored is the `ny_noaa` dataset from the `p8105.datasets` package. It contains a total of `r nrow(ny_noaa)` rows and `r ncol(ny_noaa)` columns, representing daily weather observations from various locations in New York state. The columns include: 

  * `id`: a station identifier
  * `date`: the date of the observation 
  * `prcp`: precipitation (tenths of mm)
  * `snow`: snowfall (mm)
  * `snwd`: snow depth (mm)
  * `tmax`: maximum temperature (tenths of degrees C)
  * `tmin`: minimum temperature (tenths of degrees C)

Looking at the original dataset, we see it contains substantial amount of missing data (lots of `NA` values across variables as seen from data and `summary` command). There are  `r sum(is.na(ny_noaa))` missing values in the dataset. 

### Data cleaning: 
```{r}
noaatidy_df = ny_noaa |>
  janitor::clean_names() |>
  separate(date, into = c("year", "month", "day"), convert = TRUE) |>
  mutate(
    prcp = prcp / 10 ,
    tmax = as.numeric(tmax) / 10,
    tmin = as.numeric(tmin) / 10,
    year = as.numeric(year),
    day = as.numeric(day),
    month = recode_factor(month,
          "01" = "January",
          "02" = "February",
          "03" = "March",
          "04" = "April",
          "05" = "May",
          "06" = "June",
          "07" = "July",
          "08" = "August",
          "09" = "September",
          "10" = "October",
          "11" = "November",
          "12" = "December"
    ))
```
In the data cleaning step, we cleaned the variables names, created separate variables for year, month, and day, and converted these variables to numeric variables. Additionally, variables measured in tenths of units were converted into just units (e.g., `tmin` was measured in tenths of degrees C and converted into degrees C). 

```{r}
noaatidy_df |>
  count(snow) |>
  arrange(desc(n))
```
For snowfall, the most commonly observed value is O (no recorded snowfall on most days in NY state). This makes sense because snowfall only occurs during certain times of the year and in specific conditions (i.e., there are many days with no snowfall, and a small window for possible snow). The second most commonly observed value is `NA`, suggesting missingness. 


### Making two-panel plot:
```{r}
avg_temp_df = noaatidy_df |>
  filter(month %in% c("January", "July")) |>
  group_by(id, year, month) |>
  summarize(avg_tmax = mean(tmax, na.rm = TRUE))

ggplot(avg_temp_df, aes(x = year, y = avg_tmax, group = id)) + geom_point() + geom_path() +
  facet_grid(. ~ month) +
  labs(
    title = "Average Maximum Temperature in January and July by Station",
    x = "Year",
    y = "Average Max Temperature (°C)",
  ) 
```
The two-panel plot shows the average maximum temperature in January and July across years by station. July temperatures remain stable around 25–30°C, with fewer outliers (e.g., one possible at about 14ºC), overall ranging from about 20ºC and 35ºC. Seasonal patterns are evident in January, while July shows less variation. January temperatures fluctuate more (within approximately -10 to 10ºC), with some potential outliers (e.g., lowest max temperature at about -14ºC). There does not appear to be a significant difference in average max temperatures in July or January throughout the years. The average temperatures in January (winter month) are also lower than in July (summer month). All stations appear to follow similar trends of temperature peaks and valleys within a month across the years.

### Making more two-panel plots: 
```{r}
tmax_tmin_df = noaatidy_df |>
  ggplot(aes(x = tmin, y = tmax)) +
  geom_hex() +
  labs(
    title = "Maximum vs Minimum Temperatures",
    x = "Minimum Temperature (°C)",
    y = "Maximum Temperature (°C)"
  )

tmax_tmin_df

snowfall_df = noaatidy_df |>
  filter(snow > 0, snow < 100) %>% 
  ggplot(aes(x = snow, y = as.factor(year))) +
  geom_density_ridges() +
  labs(
    title = "Distribution of Snowfall (0 < Snow < 100) by Year",
    x = "Snowfall (mm)",
    y = "Year"
  )

snowfall_df
```

The first two-panel plot (hex plot) shows the `tmax` vs `tmin` for the full dataset. The plot shows denser regions at moderate maximum temperatures (around 0ºC to 35ºC) and minimum temperatures (around -10ºC to 20ºC), suggesting most observations fall within that range. The majority of the data cluster tightly in the center of the distribution. 

The second two-panel plot shows the distribution of snowfall (values greater than 0 and less than 100) separately by year. The ridgeline plot shows a multimodal density of snowfall within a given year and distributions across years with most snowfall events concentrated around 0-35mm. There is a another group of stations that see about 45mm of snow and another group that sees nearly 80 mm. The overall patterns suggests a fairly consistent distribution of snowfall values with some variability across the years. 


## Problem 2
```{r}
covar_df = read_csv("nhanes_covar.csv", na = c("NA", ".", ""), skip = 4) |> 
  janitor::clean_names() |>
  filter(age >= 21) |>
  mutate(
    sex = case_match(
      sex, 
      1 ~ "male",
      2 ~ "female"),
    education = case_match(
      education, 
      1 ~ "less than high school",
      2 ~ "high school equivalent",
      3 ~ "more than high school"),
    age = as.numeric(age),  
    BMI = as.numeric(bmi)
  )

accel_df = read_csv("nhanes_accel.csv", na = c("NA", ".", "")) |> 
  janitor::clean_names() |>
  pivot_longer(
    min1:min1440,
    names_to = "min",
    values_to = "mims",
    names_prefix = "min"
  ) |>
  mutate(
    min = as.numeric(min)
  )

mims_df = 
  left_join(
    covar_df, accel_df, by = c("seqn"))
```
Accelerometer data was collected on 250 participants in the NHANES study. I loaded, tidied, merged, and organized two datasets to create a single `mims_df` dataset. The first dataset, `nhanes_covar`, contained participants' demographic data, and the second dataset, `nhanes_accel` contained their accelerometer data. The final dataset includes all originally observed variables, excludes participants less than 21 years of age (and those with missing demographic data), and encodes data with reasonable variable classes (i.e., not numeric, and using factors with the ordering of tables and plots in mind). The resulting dataset has `r nrow(mims_df)` rows and `r ncol(mims_df)` columns. 

```{r}
mims_df |>
  group_by(sex, education) |>
  distinct(seqn) |>
  summarize(count = n()) |>
  knitr::kable()
```
The above is a reader-friendly table for the number of men and women in each education category. We can see a summary of participants by sex (male or female) and education level (less than high school, high school equivalent, more than high school). For example, 23 females have a high school equivalent education, and 56 males have more than a high school education.

```{r}
mims_df |>
   ggplot(aes(x = education, y = age, fill = sex)) +
  geom_boxplot() + 
  labs(title = "Age Distribution by Gender and Education",
       y = "Age", 
       x = "Education") 
str(mims_df)
```
The boxplot visualization shows the age distribution for men and women in different education categories. In the "less than high school" and "high school equivalent" categories, females have a slightly higher median age compared to males. In the "more than high school" category, the median ages of men and women are quite similar (with the male median age slightly higher), both around the mid-40s. There is also considerable overlap in the age distributions of men and women across all education categories (i.e., age patterns between the sexes remain somewhat consistent regardless of education level). 

```{r}
total_activity_df = mims_df |>
  group_by(seqn, age, sex, education) |>
  summarize(total_activity = sum(mims, na.rm = TRUE))

total_activity_df |>
  ggplot(aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE) + 
  facet_wrap(~ education) +
  labs(
    title = "Total Activity vs. Age by Gender and Education Level",
    x = "Age",
    y = "Total Activity (MIMS)",
  ) 
```
I created a total activity variable for each participant, and plotted these total activities (y-axis) against age (x-axis). The plot compares men to women for each education level. Generally, across all education levels, total activity tends to decrease with age. Within each education level, females and males tend to follow similar patterns in changes in total activity as they age. Generally, females have more total activity than males, with some small exceptions (e.g., the "less than high school" education level shows males with higher total activity, in general, after the age of approximately 50). 


